Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2016-12-29 paper/ms.tex
changing the title
adding new SHO figure and editing

\title{%
Fast and scalable Gaussian process modeling of stellar variability
\vspace{-3\baselineskip}  % OMG AASTEX6 IS SO BROKEN
}

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
function is
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Dirac delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse becomes more complicated but \eq{gp-likelihood} can
still be evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number
of components in the mixture and $N$ is still the number of data points.

@@

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

@@

Another special case of the \genrp\ model of great physical interest is a
stochastically driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1-\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
}

{\item Finally, in the limit $Q \to \infty$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2-1\right]^2} \quad.
\end{eqnarray}}

\end{itemize}

% This gives a kernel with dependence
% \begin{eqnarray}\eqlabel{sho_kernel}
% %k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
% k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
% \lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
% \tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
% \end{eqnarray}
% The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
% frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
% in Figure \ref{sho_psd} for various values of $Q$.
% Note that the oscillation frequency that appears in the kernel, $d_j$, differs
% from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
% frequency, $c_j$.  This power spectrum has
% a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
% greater than zero for $Q > 1/\sqrt{2}$.

% \begin{figure}[!htbp]
% \begin{center}
% \includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
% \caption{Normalized power spectra of the stochastically-driven, damped simple
% harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
% \label{sho_psd}
% \end{center}
% \end{figure}

% This power spectrum has several limits of interest:
% \begin{itemize}
% \item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
% spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
% This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
% autocovariance function:
% \begin{equation}
% k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
% % Check to see that signs match in this equation.
% \end{equation}
% in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
% term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
% Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
% \item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
% form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
% dependence
% \begin{equation}
% k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
% \end{equation}
% corresponding to $\phi = \pi/4$.
% Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
% peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
% \item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
% The kernel approaches
% \begin{equation}
% k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
% \end{equation}
% for $Q \gg 1$.  The power spectrum approaches
% \begin{equation}
% P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
% \end{equation}
% where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
% a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
% \end{itemize}

The stochastically-driven, damped simple-harmonic oscillator can then describe
a wide range of intrinsic stellar variability, with low $Q \approx 1$
representing granulation noise, and high $Q \gg 1$ representing asteroseismic
variability.
Taking the sum over oscillators with various $Q, S_0,$ and $\omega_0$ can give
an accurate accounting of the power spectrum of stellar variability with $Q
\ge \frac{1}{2}$ and $S_0 >0$.
As this kernel is exactly described by the exponential kernel, this functional
form is tractable to solving in $\mathcal{O}(N)$ operations, as described
next.

>>>>

\title{%
Fast and scalable Gaussian process modeling with applications
to astronomical time series
\vspace{-3\baselineskip}  % OMG AASTEX6 IS SO BROKEN
}

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
function is
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Dirac delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number of
components in the mixture and $N$ is still the number of data points.

@@

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}\eqlabel{lorentz-psd}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

@@

Another special case of the \genrp\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be well approximated using
the \genrp\ framework with a small value of $f$ in \eq{approx-matern} but we
caution that this could also lead to numerical issues with the solver.
}

{\item Finally, in the limit of large $Q$, the model approaches a high
    quality oscillation with frequency $\omega_0$ and covariance function
\begin{eqnarray}
k(\tau) \approx
    \frac{1}{2}\,S_0\,\omega_0\,Q\,
    \exp\left(-\frac{\omega_0\,\tau}{2\,Q}\right)\,
    \cos\left(\omega_0\,\tau\right) \quad.
\end{eqnarray}}

\end{itemize}
\Figure{sho} shows a plot of the PSD for these limits and several other values
of $Q$.
This figure demonstrates that for $Q \le 1/2$, the model has no oscillatory
behavior and that for large $Q$, the shape of the PSD near the peak frequency
approaches a Lorentzian.

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we will take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \genrp\ method described
in the previous section.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/sho.pdf}
\caption{The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
    $S(d_j) = Q^2 = 100$.
    \figurelabel{sho}}
\end{center}
\end{figure}
